\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-06-29 analysis/paper/paper.Rmd
merge pop-stash

The splitting of these tasks gives the user the flexibility to create population-time plots with the graphics system of their choice using the object created in Step 1, while the object created in Step 2 allows one to leverage the \pkg{ggplot2} family of functions for further customization. 
1. The \code{casebase::popTime} function takes as input the original dataset along with the column names corresponding to the timescale, the event status and an exposure group of interest (optional). This will create an object of class \code{popTime} (or \code{popTimeExposure} if you specify a value for the exposure argument).  
2. The corresponding \code{plot} methods for the object created in Step 1 can be called to create the population time plot with several options for customizing the aesthetics.

To illustrate these functions we will use the European Randomized Study of Prostate Cancer Screening data which is available through the \pkg{casebase} package. The results of this study were published by [@schroder2009screening], and the dataset itself was obtained using the approach described in [@liu2014recovering].

@@

We first pass this dataset to the \code{popTime} function. If you do not specify the time and event arguments, this function will try to guess them using regular expressions. We then generate the plot by using the corresponding \code{plot} method:
```{r plot-erspc-data, eval = eval_cs1}
data("ERSPC")
ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
                       levels = c(0,1), 
                       labels = c("Control group", "Screening group"))
pt_object <- casebase::popTime(ERSPC, 
                               time = "Follow.Up.Time",
                               event = "DeadOfPrCa",
                               exposure = "ScrArm")
attr(pt_object, "exposure")
```

<!--The \code{pt_object} is a \code{data.table} containing the original dataset augmented with-->
We then pass this object to the \code{plot} function which leverages the \pkg{ggplot2} package [@ggplot2] to build population time plots. It builds the plot by sequentially adding layers, starting with a layer for the area representing the population time followed by points to show the casebase sampling mechanism. This function gives user the flexibility to add any combination of the case.series, base.series and competing events. The case series and competing events are sampled at random vertically on the plot in order to visualise the incidence density using the popTime function. That is, imagine we draw a vertical line at a specific event time. We then plot the point at a randomly sampled y-coordinate along this vertical line. This is done to avoid having all points along the upper edge of the plot (because the subjects with the least amount of observation time are plotted at the top of the y-axis). By randomly distributing them, we can get a better sense of the inicidence density. The base series is sampled horizontally on the plot using the sampleCaseBase function.

One benefit from these plots is that it allows you to see the incidence density. This can be seen from the distribution of the red dots in the above plot. We can see that more events are observed later on in time. Therefore a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on.

@@

It is important to note that the red points are random distributed across the grey area for each time of event. That is, imagine we draw a vertical line at a specific event time. We then plot the red point at a randomly sampled y-coordinate along this vertical line. This is done to avoid having all red points along the upper edge of the plot (because the subjects with the least amount of observation time are plotted at the top of the y-axis). By randomly distributing them, we can get a better sense of the inicidence density.

@@

```{r plot-stratified-erspc-data, eval = eval_cs1}
pt_object_strat <- casebase::popTime(ERSPC, 
                                     event = "DeadOfPrCa", 
                                     exposure = "ScrArm")
# class(pt_object_strat)
# pt_object
plot(pt_object_strat)
# dev.off()
# library(magrittr)
# pt <- plot(pt_object_strat)
# pt %>% class

# library(ggplot2)
# pt + theme_minimal()

```

>>>>

1. The \code{casebase::popTime} function takes as input the original dataset along with the column names corresponding to the timescale, the event status and an exposure group of interest (optional). This will create an object of class \code{popTime} (or \code{popTimeExposure} if you specify a value for the exposure argument).  
2. The corresponding \code{plot} methods for the object created in Step 1 can be called to create the population time plot with several options for customizing the aesthetics.

The splitting of these tasks gives the user the flexibility to create population-time plots with the graphics system of their choice using the object created in Step 1, while the object created in Step 2 allows one to leverage the \pkg{ggplot2} [@ggplot2] family of functions for further customization.

To illustrate these functions we will use the European Randomized Study of Prostate Cancer Screening (ERSPC) data which is available through the \pkg{casebase} package. The results of this study were published by [@schroder2009screening], and the dataset itself was obtained using the approach described in [@liu2014recovering].

@@

```{r plot-erspc-data, eval = eval_cs1, echo = TRUE}
data("ERSPC")
ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
                       levels = c(0,1), 
                       labels = c("Control group", "Screening group"))

pt_object <- casebase::popTime(ERSPC, 
                               time = "Follow.Up.Time",
                               event = "DeadOfPrCa",
                               exposure = "ScrArm")
attr(pt_object, "exposure")

pt_object_strat <- casebase::popTime(ERSPC, 
                                     time = "Follow.Up.Time",
                                     event = "DeadOfPrCa",
                                     exposure = "ScrArm")
attr(pt_object, "exposure")
```

<!--The \code{pt_object} is a \code{data.table} containing the original dataset augmented with-->
We then pass this object to the corresponding \code{plot} method which leverages the \pkg{ggplot2} package to build population time plots.

```{r plot-stratified-erspc-data, eval = eval_cs1, echo = TRUE}
pacman::p_load(cowplot)
plot(pt_object, add.case.series = FALSE) + cowplot::theme_minimal_hgrid(12, rel_small = 1)
theme_cowplot
plot(pt_object, legend = TRUE, theme.params = list(legend.position = "bottom"))

# plot(pt_object, add.base.series = TRUE, ratio = 1, legend = TRUE)
```

Figure~\ref{fig:plot-stratified-erspc-data} is built sequentially by adding layers for the area representing the population time in grey, followed by points to show the case series. The case series are sampled at random vertically on the plot in order to visualise the incidence density using the \code{popTime} function. That is, imagine we draw a vertical line at a specific event time. We then plot the point at a randomly sampled y-coordinate along this vertical line. This is done to avoid having all points along the upper edge of the plot because the subjects with the least amount of observation time are plotted at the top of the y-axis. By randomly distributing the points, we can get a better sense of the inicidence density. In the ERSPC data, we can see that more events are observed later on in time. Therefore a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on.

casebase sampling mechanism. This function gives user the flexibility to add any combination of the case series, base series and competing events.  The base series is sampled horizontally on the plot using the sampleCaseBase function.

@@

@@

### 2020-06-29 analysis/paper/paper.tex
merge pop-stash

To illustrate these functions we will use the European Randomized Study
of Prostate Cancer Screening data which is available through the
\pkg{casebase} package. The results of this study were published by
\citep{schroder2009screening}, and the dataset itself was obtained using
the approach described in \citep{liu2014recovering}.

@@

@@

@@

\begin{center}\includegraphics{../figures/unnamed-chunk-2-1} \end{center}

>>>>

The splitting of these tasks gives the user the flexibility to create
population-time plots with the graphics system of their choice using the
object created in Step 1, while the object created in Step 2 allows one
to leverage the \pkg{ggplot2} family of functions for further
customization. To illustrate these functions we will use the European
Randomized Study of Prostate Cancer Screening data which is available
through the \pkg{casebase} package. The results of this study were
published by \citep{schroder2009screening}, and the dataset itself was
obtained using the approach described in \citep{liu2014recovering}.

@@

We first pass this dataset to the \code{popTime} function. If you do not
specify the time and event arguments, this function will try to guess
them using regular expressions. We then generate the plot by using the
corresponding \code{plot} method:

@@

The randomization of the Finnish cohorts were carried out on January 1
of each of the 4 years 1996 to 1999. This, coupled with the uniform
December 31 2006 censoring date, lead to large numbers of men with
exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time
(i.e.~from right to left), the population-time plot shows the
recruitment pattern from its beginning in 1991, and in particular the
Jan 1 entries in successive years. Tracked forwards in time (i.e.~from
left to right), the plot for the first three years shows attrition due
entirely to death (mainly from other causes). Since the Swedish and
Belgian centres were the last to close their recruitment - in December
2003 - the minimum potential follow-up is three years. Tracked further
forwards in time (i.e.~after year 3) the attrition is We can also create
exposure stratified plots by specifying the \code{exposure} argument in
the \code{popTime} function:

@@

\begin{center}\includegraphics{../figures/unnamed-chunk-1-1} \end{center}
